### 2025-04-09 correlation.qmd
20250409 - nonlinear correlation, robust correlation, outliers, restricted range

```{r}
library("petersenlab")
library("XICOR")
library("psych")
library("DescTools")
library("psych")
library("corrplot")
library("tidyverse")
```

@@

@@

@@

@@

>>>>

```{r}
library("petersenlab")
library("XICOR")
library("psych")
library("DescTools")
library("psych")
library("corrplot")
library("correlation")
library("tidyverse")
```

@@

An example of a nonlinear correlation is a distance correlation and the xi ($\xi$) coefficient.
These estimates of nonlinear correlation only range from 0–1 (where 0 represents no association and 1 represents a strong association), so they are interpreted differently than a traditional correlation coefficient (which can be negative).
We can estimate a distance correlation using the `correlation()` function of the correlation package [@R-correlation; @Makowski2020].

```{r}
correlation(
  data = player_stats_seasonal %>% select(draftround, fantasyPoints),
  method = "distance"
)
```

@@

The correlation coefficient is strongly impacted by outliers—i.e., extreme (i.e., very large or very small) values relative to the other observations.
Consider the following example:

```{r}
set.seed(52242)

v1 <- rnorm(10)
v2 <- rnorm(10)

cor.test(
  v1,
  v2
)
```

The associated scatterplot and best-fit line is in @fig-plotWithoutOutliers.

```{r}
#| label: fig-plotWithoutOutliers
#| fig-cap: "Scatterplot Without Strong Outliers."
#| fig-alt: "Scatterplot Without Strong Outliers."
#| code-fold: true

plot(
  v1,
  v2
)

petersenlab::addText(
  v1,
  v2,
  ycoord = 0.7
)

abline(lm(
  v2 ~ v1),
  col = "black")
```

Now, let's add one outlier.

```{r}
v1[length(v1) + 1] <- 4
v2[length(v2) + 1] <- 4

cor.test(
  v1,
  v2
)
```

The associated scatterplot and best-fit line for the updated data are in @fig-plotWithOutlier.

```{r}
#| label: fig-plotWithOutlier
#| fig-cap: "Scatterplot With Outlier."
#| fig-alt: "Scatterplot With Outlier."
#| code-fold: true

plot(
  v1,
  v2
)

petersenlab::addText(
  v1,
  v2
)

points(
  v1[length(v1)],
  v2[length(v2)],
  pch = 19,
  col = "red"
)

text(
  x = 3.8,
  y = 3.8,
  labels = "outlier")

abline(lm(
  v2 ~ v1),
  col = "black")
```

Note how the association was not close to being statistically significant with the outlier, but with the outlier added, the association is statistically significant.
One way to combat this is to use methods for estimating the association between variables that are robust to (i.e., less impacted by) outliers.
In the next section, we describe robust correlation methods.

@@

There are various approaches to estimating correlation in the presence of outliers—so-called *robust correlation* methods.

One approach is the biweight midcorrelation, which is based on the median rather than the mean, and is thus less sensitive to outliers.
Another approach is the percentage bend correlation, which gives less weight to observations that are farther away from the median.
We can estimate the each using the `correlation()` function of the correlation package [@R-correlation; @Makowski2020].

```{r}
correlation(
  data = data.frame(v1, v2),
  method = "biweight"
)

correlation(
  data = data.frame(v1, v2),
  method = "percentage"
)
```

@@

In addition to being [impacted by outliers](#sec-correlationOutliers), correlations can also be greatly impacted by restricted variability or restricted range [@Petersen2024a; @PetersenPrinciplesPsychAssessment].
Correlation depends on variability; if there is no or limited variability, it can be difficult to detect an association with another variable.
Thus, if a variable has restricted range—such as owing to a floor effect or ceiling effect—that tends to artificially weaken associations.
For instance, a floor effect is one in which many of the scores are the lowest possible score.
By contrast, a ceiling effect is one in which many of the scores are the highest possible score.

Consider the following association between air yards share and receiving yards:

```{r}
cor.test(
  player_stats_seasonal$air_yards_share,
  player_stats_seasonal$receiving_yards
)
```

There is a statistically significant, moderate positive association between air yards share and receiving yards, as depicted in @fig-scatterplotWithoutRangeRestriction.

```{r}
#| label: fig-scatterplotWithoutRangeRestriction
#| fig-cap: "Scatterplot of Player Receiving Yards (Season) by Air Yard Share. The best-fit line is in black."
#| fig-alt: "Scatterplot of Player Receiving Yards (Season) by Air Yard Share. The best-fit line is in black."
#| code-fold: true

plot_scatterplotWithoutRangeRestriction <- ggplot2::ggplot(
  data = player_stats_seasonal %>% filter(air_yards_share > 0),
  aes(
    x = air_yards_share,
    y = receiving_yards)) +
  geom_point(
    aes(
      text = player_display_name, # add player name for mouse over tooltip
      label = season), # add season for mouse over tooltip
    alpha = 0.10) +
  geom_smooth(
    method = "lm",
    color = "black") +
  #geom_smooth() +
  coord_cartesian(
    ylim = c(0,NA),
    expand = FALSE) +
  labs(
    x = "Air Yards Share",
    y = "Receiving Yards",
    title = "Receiving Yards (Season) by Air Yards Share",
    subtitle = "(Among Players with Air Yards Share > 0)"
  ) +
  theme_classic()

plotly::ggplotly(plot_scatterplotWithoutRangeRestriction)
```

Now, consider the same association when we restrict the range to examine only those players whose air yards share was more than 25 (thus resulting in less variability):

```{r}
cor.test(
  player_stats_seasonal %>% 
    filter(air_yards_share > 25) %>% 
    select(air_yards_share) %>% 
    pull(),
  player_stats_seasonal %>% 
    filter(air_yards_share > 25) %>% 
    select(receiving_yards) %>% 
    pull()
)
```

The association is no longer statistically significant, and the association is weak in terms of [effect size](#sec-effectSizeThresholds), as depicted in @fig-scatterplotWithRangeRestriction.

```{r}
#| label: fig-scatterplotWithRangeRestriction
#| fig-cap: "Scatterplot of Player Receiving Yards (Season) by Target Share. The linear best-fit line is in black. The nonlinear best-fit line is in blue."
#| fig-alt: "Scatterplot of Player Receiving Yards (Season) by Target Share. The linear best-fit line is in black. The nonlinear best-fit line is in blue."
#| code-fold: true

plot_scatterplotWithRangeRestriction <- ggplot2::ggplot(
  data = player_stats_seasonal %>% filter(air_yards_share > 25),
  aes(
    x = air_yards_share,
    y = receiving_yards)) +
  geom_point(
    aes(
      text = player_display_name, # add player name for mouse over tooltip
      label = season)) + # add season for mouse over tooltip)
  geom_smooth(
    method = "lm",
    color = "black") +
  #geom_smooth() +
  coord_cartesian(
    ylim = c(0,NA),
    expand = FALSE) +
  labs(
    x = "Air Yards Share",
    y = "Receiving Yards",
    title = "Receiving Yards (Season) by Air Yards Share",
    subtitle = "(Among Players with Air Yards Share > 25)"
  ) +
  theme_classic()

plotly::ggplotly(plot_scatterplotWithRangeRestriction)
```

### 2025-04-09 index.qmd
20250409 - nonlinear correlation, robust correlation, outliers, restricted range

# Automatically create a bib database for R packages
packagesForBibliography <- c(
  "petersenlab", "ffanalytics",
  "knitr", "rmarkdown", "dplyr", "ggplot2", "tidyverse", "downlit", "xml2",
  "nflreadr", "nflplotR", "nflfastR", "nfl4th", "gsisdecoder", "progressr",
  "DescTools", "pwr", "pwrss", "WebPower", "XICOR", "dagitty", "ggdag",
  "lubridate", "ggtext", "gghighlight", "ggExtra", "grid", "patchwork", "pROC",
  "lme4", "lmerTest", "MuMIn", "emmeans", "pbkrtest", "sjstats", "AICcmodavg",
  "rstan", "brms", "bbmle", "fitdistrplus", "mclust", "magrittr", "viridis",
  "plotly", "quantmod", "fPortfolio", "xts", "zoo", "forecast", "stringi", #"webshot2", 
  "parallel", "parallelly", "doParallel", "missRanger", "ggridges", "powerjoin",
  "caret", "gpboost", "cmdstanr", "learnr", "remotes", "corrplot" #"mgcv",
)

>>>>

# Automatically create a bib database for R packages
packagesForBibliography <- c(
  "petersenlab", "ffanalytics",
  "knitr", "rmarkdown", "dplyr", "ggplot2", "tidyverse", "downlit", "xml2",
  "nflreadr", "nflplotR", "nflfastR", "nfl4th", "gsisdecoder", "progressr",
  "DescTools", "pwr", "pwrss", "WebPower", "XICOR", "dagitty", "ggdag",
  "lubridate", "ggtext", "gghighlight", "ggExtra", "grid", "patchwork", "pROC",
  "lme4", "lmerTest", "MuMIn", "emmeans", "pbkrtest", "sjstats", "AICcmodavg",
  "rstan", "brms", "bbmle", "fitdistrplus", "mclust", "magrittr", "viridis",
  "plotly", "quantmod", "fPortfolio", "xts", "zoo", "forecast", "stringi", #"webshot2", 
  "parallel", "parallelly", "doParallel", "missRanger", "ggridges", "powerjoin",
  "caret", "gpboost", "cmdstanr", "learnr", "remotes", "corrplot", "correlation" #"mgcv",
)
